Network Data Model and BerlinMOD Benchmark

### 2010-04-15 Algebras/Network/BerlinMODNetwork/paper/PaperBerlinMODAndNetworkDataModel.tex
paper updated.

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[boxed]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption,subcaption}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
\newcommand{\secver}{2.9.StableNewFlob}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past several data models for the representation of spatio-temporal data
objects have been developed. Among others we can categorize this data models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In the mean time two representants
of this data models, one for each category, have been implemented in
\secondo{} DBMS. We used the \bmodb{} to compare the capabilities of the both
data models. The results are shown in this paper.

The very good results of the network data model in this comparsion encourages us
to propose a extension of the \bmodb{} with a additional query set BerlinMOD/Net.
This new query set covers the specialised challenges for network constrained
data models and enables us to use the \bmodb{} to compare the capabilities of
different network data models with respect to this specialised network challenges.
\end{abstract}
\section{Introduction}
In the past several data models for the representation of spatial and
spatio-temporal data objects have been developed. Among others we can categorize
them into data models for objects moving freely in (two dimensional) space and
data models for network constrained moving objects. For both categories
several different data models have been presented like
\cite{335426,chenzaniolosqlst} for spatio-temporal data objects moving freely in
 space and \cite{1146465,956692,VazWolfNetMod} for spatio-temporal data objects
that are constrained by given networks to name just a few.

For our experiments we choosed one example data model as representative for
each of the both categories to compare the capabilities of the both different
data models. Spatio-temporal objects moving freely in two dimensional space are
represented by the data model presented in \cite{335426}. And network constrained
data models are represented by the network data model presented in \cite{1146465}.
Both data models are available in the \secondo{} DBMS and use the same temporal representation. So we can exclude that the use of different DBMS or temporal
representations bias the results of our comparision.

As we use \secondo{} it seems to be a logical next step to use the \bmodb{} \cite{BerlinMODVLDB} to compare the capabilities of the
both data models. The \bmodb{} is best to our knowledge the first benchmark for
complete spatio-temporal database systems and it is also available in the \secondo{}
 DBMS. Furthermore the \bmodb{} data model is the data model of freely movement
in two dimensional space that we use for our comparison. So we only have to
transform the \bmodb{} spatial and spatio-temporal data types once into our
network data model representation. This avoids sources of errors and makes the
correctness controll of the query results more easy.

In our experiments the network data model outperforms the data model of free
movement in space significantly. The total run time for all benchmark queries
is less than half as long for the network model as the total run time for
the data model of free movement in the space. Such that it seems useful to
develope specialised databases and operations for network constrained moving
objects. In the same way the storage space used for the trips in the network
data model is less than 60\% of the storage space used by the data model of free
movement in the space. Also the number of units in the network data model is
less than 50\% of the number of units in the data model of free movement in the
space.

The very good results of the network data model encourages us to propose the extension
of the \bmodb{} by a query set BerlinMOD/Net. The new query set covers the specialised
challenges of network data models like network distance computing which are not
covered by the original \bmodb{}. The new query set should enable us to compare
the capabilities of different network data models with respect to the special
challenges of network constrained data models.

The rest of the paper is organised as follows: In section \ref{sec:relWork} we give a
short reminder of the underlying \secondo{} DBMS (\ref{sec:secondo}), the \bmodb{} (\ref{sec:bmodb}) and the two data models (\ref{sec:bmodbdatamod}
 and \ref{sec:netdatamod}) we choosed as representatives for our comparision.
In section \ref{sec:bmodbNetDataMod} we present the setup for our experiments (\ref{sec:scenario}) and describe our transformation of the \bmodb{} data and
queries in the network data model representation (\ref{sec:Translation}). The
results of our experiments are presented in section (\ref{sec:results}). At least
we present our proposed extension of the \bmodb{} for network data models in section \ref{sec:newqueries} and conclude our work in section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
As mentioned before in the past other data models for free movement in the space \cite{335426,chenzaniolosqlst} and for network constrained movement have been
presented \cite{1146465,956692,VazWolfNetMod}. As well there are some more
benchmarks \cite{COSTBenchmark, QueriesTheodoridis} and database systems for
spatial and spatio-temporal datatypes \cite{HERMES,1054151}. But they don't
provide a combination of different implemented and supported data types together
 with a existing benchmark like the actual \secondo{} DBMS. So it is self-evident
for us to use the \secondo{} DBMS in combination with the provided data types and
the \bmodb{} to compare the capabilities of the both different data models.

In the next subsections we give short reminders of the \secondo{} database system (\ref{sec:secondo}), the \bmodb{} (\ref{sec:bmodb}), and the both data models (\ref{sec:bmodbdatamod}, \ref{sec:netdatamod}) we used in our experiments.
\subsection{Secondo}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in \cite{686903,1054151} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
 single data models. Hence \secondo{} can be easily extended by the user
implementing so called \secondo{} algebra modules to introduce  new data types
and operations on this data types. The user also may define additional viewers
for the graphical user interface or write optimization rules or cost functions
to extend the optimizer. \secondo{} is free available in the web \cite{secondoweb}
and comes with a number of already implemented spatial and spatio-temporal data types
and operations including the spatio-temporal data model of free movement in the space \ref{sec:bmodbdatamod} and the network data model \ref{sec:netdatamod}. Furthermore the \bmodb{} descripted in \ref{sec:bmodb} has been developed in the \secondo{} DBMS.
For our experiments we used the \secondo{} version \secver{}.
\subsection{BerlinMOD Benchmark}
\label{sec:bmodb}
The \bmodb{} was presented in \cite{BerlinMODVLDB} \nocite{BerlinMOD} and the
provided scripts for the data generator are implemented as \secondo{} DBMS operations.
The \bmodb{} is available in the web \cite{berlinmodweb} and provides a well defined
data-set and queries for the experimental evaluation of different moving object
data representations. The \bmodb{} emphasises the development of complete systems
and simplifies experimental repeatability pointing out the weakness and the potency
of the benchmarked systems.

The data-sets of the \bmodb{} are created using the street map of the German
captial Berlin \cite{bbike} and statistical data about the regions of Berlin
\cite{bevberlin,berlinstadtatlas} as input relations.
The created moving objects represent cars driving in the streets of Berlin,
simulating the behaviour of people living and working in Berlin.
Every moving object has a home node and a work node. Every weekday each car will
do a trip from the home node to work the work node in the morning and vice versa
in the late afternoon. Besides this randomly choosen cars will make additional
trips in the evening and up to six times at the weekend to randomly choosen
targets in Berlin. Because for all this the street network of Berlin is used to
generate the moving object data sets the \bmodb{} can also be used to benchmark
network constrained moving object data models.

The number of observed cars and the duration of the observation period can be
influenced setting the $SCALEFACTOR$ to different values in the data generation
script of the \bmodb{}. For example at $SCALEFACTOR$ 1.0 the data generator
creates 2000 moving objects observed for 28 days. Each of them sending a
GPS-signal every 2 seconds. This simulated signals are simplified such that time
intervals when a car doesn't move or moves in the same direction at the same
speed are merged into one moving unit. This means that if for example a car parks
8 hours in front of the work node there will be only one entry in the cars history of movement with a time interval of 8 hours instead of 14.400 entries one for each
GPS interval.

@@

In the OBA the complete history for each moving object
is kept together into one single entry. There is only one relation (dataScar)
containing one tuple for each object consisting of the spatio-temporal data of
the object (journey), the licence, the type, and the model of the object.

In the TBA we have two relations. One of them (dataMcar) contains the static data
for each object like licence, type, and model together with an object identifier.
The other relation (dataMtrip) contains for each object identifier several tuples
each of them containing a single trip of the moving object (e.g. each time the car
drives from home node to work node) or a longer stop (e.g. the time the car parks
in front of the office).

Besides the moving objects the \bmodb{} provides sets of $QueryPoints$,
$QueryRegions$, $QueryInstants$, $QueryPeriods$, and $QueryLicences$,
each of them containing 100 pseudo randomly generated data objects (points, regions,
time instants, time intervals, and licences) used in the benchmark queries.

The \bmodb{} provides two sets of queries. One set BerlinMOD/R addresses
range queries and the  other one BerlinMOD/NN nearest neighbour queries.
We will focus on the range queries in this paper, which are the main aspect of the
\bmodb{} up to now.

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects:
\begin{itemize}
  \item object identity (known / unknown),
  \item dimension (standard / spatial / temporal / spatio-temporal),
  \item query interval (point / range / unbounded),
  \item condition type (single object / object relations),
  \item aggregation (with or without aggregation).
\end{itemize}
We will present the 17 queries in more detail in section \ref{sec:queries}
together with our network data model translation of this queries.
\subsection{BerlinMOD Data Model}
\label{sec:bmodbdatamod}
The data model used by the \bmodb{} is the same data model of freely moving in
two dimensional space presented in \cite{594784,335426,352963}. All spatial
positions are given in x,y-coordinates. A single spatial position is represented
by the data type \dt{point}. A \dt{point} consists of a pair of \dt{real} values
interpreted as x,y-coordinates in the two dimensional space.
Streets are represented by \dt{line} values.
A \dt{line} value consists of a set of half segments representing the geometry
of the line in the two dimensional space. Each half segment consists of two
\dt{point} values
which are interpreted as start and end point of the half segment. Regions are
represented by the data type \dt{region}. A \dt{region} consists of a set of
half segments interpreted as outer (and inner) border of the region in the two
dimensional space.

All this spatial data types and many standard data types can be lifted to become
time dependent \dt{moving} values. For all data types \dt{$\alpha$} the constructor
\dt{moving} creates a new data type \dt{moving}(\dt{$\alpha$}) (short form \dt{m$\alpha$}).
A car may be represented by a \dt{mpoint}. A \dt{mpoint} is a point object changing
its position within time. Therefore a \dt{mpoint} consist of a set of units called
\dt{unit}(\dt{point}) (short form \dt{upoint}). Each \dt{upoint} consists of a time
interval and two \dt{point} values. The first point represents the position of the
\dt{mpoint} at the start of the time interval and the second point represents the
position of the \dt{mpoint} at the end of the time interval. It is assumed that
the point moves on the straight line between this two points with constant speed.
The speed is given by the ratio from the distance of the two points and the length
of the time interval of the unit. All units of a \dt{mpoint} must have disjoint
time intervals, because a car cannot be at two different positions at the same time.
The units are sorted by ascending time intervals. This spatio-temporal data model of \dt{moving} allows us to compute the position of a \dt{mpoint} at every time instant
within its definition time. We can also compute the time instant the point passed a
given position assumed the \dt{mpoint} ever passes this position. The position of a
\dt{point} at a given time instant is represented by a \dt{intime}(\dt{point}) (short form \dt{ipoint}). A \dt{ipoint} consists of a time instant and a \dt{point} value.

Some other data types of \secondo{} which are used in the \bmodb{} are shown in table \ref{tab:bmodbdatatypes}.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{\textwidth}{|l|X|}
\hline
\textbf{Data Type} & \textbf{Description} \\
\hline
\dt{bool} & Usual boolean data type.\\
\hline
\dt{int} & Usual integer number.\\
\hline
\dt{real} & Usual real number.\\
\hline
\dt{instant} & A point in time.\\
\hline
\dt{periods} & A set of disjoint and not connected time intervals.\\
\hline
\dt{mbool} & A time dependent boolean value. The value within each \dt{ubool} will
 be constant \true{} or \false{} \\
\hline
\dt{mreal} & Time dependent real number. Each unit will be defined by a function
of time representing the \dt{real} value at each time instant.\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Other Data Types of \bmodb{}}
\label{tab:bmodbdatatypes}
\end{center}
\end{table}
\subsection{Network Data Model}
\label{sec:netdatamod}
The central idea of the network data model presented in \cite{1146465} is that
every movement is constrained by a network and every position is given related
to this network. The data type \dt{network} is the central data type in the network
data model. All other data types of the network data model are related to a
given \dt{network} object by the unique network identifier that is part of
each \dt{network} object.

The \dt{network} object contains all spatial information of the represented network
in three main relations. The first relation contains the attributes of the routes
(streets) like id, route curve, route length, and two boolean flags indicating
if the route starts at the lexicographically smaller end point or not and if the
lanes of the route are separated like on German Highways or not.
The second relation
contains all attributes of the junctions (street crossings) like the identifiers
of the first and second route crossing in the junction, the distance of the
junction from the start of the first respectively second route, tuple identifiers
of the both routes in the routes relation, tuple identifiers of the sections
connected by this junction in the sections relation, and a connectivity code
telling us which lanes of the two routes are connected by the junction. The third
main relation of the \dt{network} object is the sections relation containing the
attributes of the sections (street parts between two junctions or a junction and
the end of the street) like the route identifier of the route the section belongs to,
the tuple identifier of this route in the routes relation, start and end position
of the section on the route, section curve, and two boolean flags $startssmaller$
and $dual$ with the same meaning as in the routes relation.

We introduce four B-Tree indexes and one R-Tree index to support faster query execution.
The four B-Trees indicate the route identifier attributes in the three main relations.
And the R-Tree indicates the curve attribute of the routes relation. Furthermore
there are two sets which provide a fast access from each section to their adjacent
sections with respect to the driving direction. Two sections are adjacent if their
lanes are connected by a junction.

Single positions in the network are given as \dt{gpoint} values. Besides the
network identifier a \dt{gpoint} consists of a route identifier, a distance from
the start of the route to the position of the \dt{gpoint} and a $side$ value
({$up$, $down$, $none$). The $side$ value is necessary if the route is dual.
It tells us if the position
is reachable from the $up$ or the $down$ side of the route. For simple
routes or positions which are reachable from both sides of the route the side
value is always $none$.

Parts of the network, regardless if they represent paths or regions, are given as
\dt{gline} values. Besides the network identifier a \dt{gline} consists of a set
of route intervals, and two boolean flags telling us if the \dt{gline} is defined
and if the set of route intervals is sorted.

Each route interval consists of a route identifier identifying the route
the route interval belongs to, and the start and the end position from the
route interval on this route
\footnote{In the original paper the
route interval includes a additional parameter $side$ analogous to the \dt{gpoint}.
But this parameter is not part of the implementation yet.}.
We call a \label{sec:sortedgline}
set of route intervals sorted if the following conditions are fullfilled:
\begin{itemize}
	\item all route intervals are disjoint
	\item the route intervals are stored in ascending order of their route identifiers
	\item if two disjoint route intervals have the same route identifier the route interval
  with the smaller start position is stored first
	\item for all route intervals in the set the condition: $startPosition \le endPosition$ holds
\end{itemize}
We introduced the sorted \dt{gline} because many algorithms take profit from sorted
\dt{gline} values. For example the computation if a \dt{gpoint} is inside a
\dt{gline} needs O($n$) time for unsorted and O($\log n$) time for
sorted \dt{gline} values, if $n$ is the number of the route intervals in the
\dt{gline}.

@@

Mostly similar to the \dt{mpoint} of the other data model we implemented a
\dt{mgpoint}. A \dt{mgpoint} consists of a set of \dt{ugpoint} with disjoint
time intervals. Each \dt{ugpoint} consists of a time interval and two \dt{gpoint}
values. Every time the \dt{mgpoint} changes the route or the speed a new
\dt{ugpoint} is written. Each \dt{ugpoint} is assumed to follow the same route
from the start to the end position at the same speed. So accordingly to the
\dt{mpoint} we can compute the network position of the \dt{mgpoint} at ever time
instant within the definition time of the \dt{mgpoint} as \dt{intime}(\dt{gpoint}).

>>>>

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[boxed]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption,subcaption}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
\newcommand{\secver}{2.9.StableNewFlob}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past several data models for the representation of histories of  spatio-temporal data
objects have been developed. Among others we can categorize these data models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. We selected two representants,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compared their capabilities with the \bmodb{}. In this paper we present our
translation from the \bmodb{} into the network constrained data model and show
that the network constrained data model outperforms the data model of free movement
in the two dimensional space by orders of magnitudes.

%The very good results of the network data model in this comparsion encourages us
%to propose a extension of the \bmodb{} with a additional query set BerlinMOD/Net.
%This new query set covers the specialised challenges for network constrained
%data models and enables us to use the \bmodb{} to compare the capabilities of
%different network data models with respect to this specialised network challenges.
\end{abstract}
\section{Introduction}
In the past several data models for the representation of spatial and
spatio-temporal data objects have been developed. Among others we can categorize
them into data models for objects moving freely in two dimensional space and
data models for network constrained moving objects. For both categories
several different data models have been presented like
\cite{335426,chenzaniolosqlst} for spatio-temporal data objects moving freely in
two dimensional space and \cite{1146465,956692,VazWolfNetMod} for
spatio-temporal data objects that are constrained by given networks,
to name just a few. Objects which are restricted to use existing networks, like
cars to road networks, can be represented as moving point object in both data
models, while objects, which are not restricted by a given network, like people,
can be represented as moving point object only in the data models of free movement
in space. So, why do we spend time in developing network data models, if the network
constrained moving objects can be represented elsewhere in the data model of free
movement in the space?

We decided to compare the capablities of the both different data models. For our
experiments we choosed one example data model as representative for
each of the both categories. Spatio-temporal objects moving freely in the two
dimensional space are represented by the data model presented in \cite{335426}.
And network constrained data models are represented by the network data model
presented in \cite{1146465}. We choosed this two representatives, because both
data models are available in the \secondo{} DBMS and use the same temporal
representation. So we can exclude that the use of different DBMS or temporal
representations bias the results of our comparision.

As we use \secondo{} it seems to be a logical next step to use the \bmodb{}
\cite{BerlinMODVLDB} to compare the capabilities of the
both data models. The \bmodb{} is best to our knowledge the first benchmark for
complete spatio-temporal database systems and it is also available in the \secondo{}
DBMS. Furthermore the \bmodb{} data model is the data model of freely movement
in two dimensional space that we use for our comparison. So we only have to
transform the \bmodb{} spatial and spatio-temporal data types once into our
network data model representation. This avoids sources of errors in transformation
and makes it more easy to control the query results for correctness. Furthermore
the data generated by the \bmodb{} data generator is restricted to the streets of the
German capital Berlin, such that it can be translated into a network data model
environment.

In this paper we describe the transformation of the \bmodb{} benchmark data and
queries into network data model representation. And show that the network data
model outperforms the data model of free movement in space significantly in our
experiments. The network data model uses less than 60\% of the storage space of
the data model of free movement in space, mostly caused by the fact that the number
of units for the moving point objects in the network data model is less than 50\% of
the number of units in the data model of free movement in space. The total run times
for the \bmodb{} of the network data model is less than 50\% of the total run time
of the data model of free movement in space. We think that this results show that
it is useful to develope specialised data models for network constrained moving objects
to save storage space and reduce query run times.

%The very good results of the network data model encourages us to propose the extension
%of the \bmodb{} by a query set BerlinMOD/Net. The new query set covers the specialised
%challenges of network data models like network distance computing which are not
%covered by the original \bmodb{}. The new query set should enable us to compare
%the capabilities of different network data models with respect to the special
%challenges of network constrained data models.

The rest of the paper is organised as follows: In section \ref{sec:relWork} we give a
short reminder of the underlying \secondo{} DBMS (\ref{sec:secondo}), the \bmodb{}
(\ref{sec:bmodb}) and the two data models (\ref{sec:bmodbdatamod}
and \ref{sec:netdatamod}) we choosed as representatives for our comparision.
The translation of the \bmodb{} data and query set in the network data model
representation is described in section (\ref{sec:Translation}). The resulting
experimental benchmark setup is described in section \ref{sec:scenario} followed
by the results of our experiments in section (\ref{sec:results}).
%At least we present our proposed extension of the \bmodb{} for network data models in
%section \ref{sec:newqueries}.
We conclude our work in section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
In the past other data models for free movement in the space
\cite{335426,chenzaniolosqlst} and for network constrained movement have been
presented \cite{1146465,956692,VazWolfNetMod}. As well there are some more
benchmarks \cite{COSTBenchmark, QueriesTheodoridis} and database systems for
spatial and spatio-temporal datatypes \cite{HERMES,1054151}. But best to our
knowledge they don't actually provide a combination of different implemented
and supported data models together with a existing benchmark feasible for moving
objects in space and network constrained moving objects like the actual
\secondo{} DBMS. So it is self-evident for us to use the \secondo{} DBMS in
combination with the provided data models and the \bmodb{} to compare the
capabilities of the network constrained data model and the data model of free
movement in spache provided with the \secondo{} DBMS.

In the next subsections we give short reminders of the \secondo{} database system
(\ref{sec:secondo}), the \bmodb{} (\ref{sec:bmodb}), and the both data models
(\ref{sec:bmodbdatamod}, \ref{sec:netdatamod}) we used in our experiments.
\subsection{Secondo}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in \cite{686903,1054151} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
single data models. Hence \secondo{} can be easily extended by the user
implementing algebra modules to introduce  new data types and operations on
this data types. The user may define additional viewers for the graphical user
interface or write additional optimization rules or cost functions to extend the
the optimizer. Since \secondo{} version 2.9 the user may publish his extensions
as \secondo{} plugin such that other users can use this plugin to extend their own
\secondo{} system to use the functionalities or repeat the published experiments.
\secondo{} is free available in the web \cite{secondoweb} and comes with a number
of already implemented spatial and spatio-temporal data types
and operations including the spatio-temporal data model of free movement in the
two dimensional space \ref{sec:bmodbdatamod} and the network data model
\ref{sec:netdatamod} used in our work. Furthermore the \bmodb{} described in
\ref{sec:bmodb} has been developed in the \secondo{} DBMS. For our experiments
we used the \secondo{} version \secver{}.
\subsection{BerlinMOD Benchmark}
\label{sec:bmodb}
The \bmodb{} was presented in \cite{BerlinMODVLDB} \nocite{BerlinMOD} and the
provided scripts for the data generator are implemented as \secondo{} DBMS operations.
The \bmodb{} is available in the web \cite{berlinmodweb} and provides a well defined
data-set and queries for the experimental evaluation of the capabilities of
spatial and spatio-temporal database systems dealing with histories of moving
objects. The \bmodb{} emphasises the development of complete systems
and simplifies experimental repeatability pointing out the weakness and the potencies
of the benchmarked systems.

The data-sets of the \bmodb{} are created using the street map of the German
captial Berlin \cite{bbike} and statistical data about the regions of Berlin
\cite{bevberlin,berlinstadtatlas} as input relations.
The created moving objects represent cars driving in the streets of Berlin,
simulating the behaviour of people living and working in Berlin.
Every moving object has a home node and a work node. Every weekday each car will
do a trip from the home node to work the work node in the morning and vice versa
in the late afternoon. Besides this randomly choosen cars will make additional
trips in the evening and up to six times at the weekend to randomly choosen
targets in Berlin. The \bmodb{} uses the data model of free movement in two
dimensional space described in section \ref{sec:bmodbdatamod}. Because the \bmodb{}
generates all data sets restricted to the street map of Berlin the \bmodb{} can
also be used for network constrained data models, if the spatial and spatio-temporal
data types are translated into a corresponding network data model, like we did
for our experiments.

The number of observed cars and the duration of the observation period can be
influenced by the user setting the $SCALEFACTOR$ to different values in the data
generation script of the \bmodb{}. For example at $SCALEFACTOR$ 1.0 the data generator
creates 2000 moving point objects observed for 28 days. Each of them sending a
GPS-signal every 2 seconds. This simulated signals are simplified such that time
intervals when a car doesn't move or moves in the same direction at the same
speed are merged into one moving unit. This means that, if a car stays in front
of the work node for 8 hours there will be only one entry in the cars
history of movement with a time interval of 8 hours instead of 14.400 entries
one for each GPS interval.

@@

In the OBA the complete history for each moving object is kept together into
one single entry. There is only one relation $dataScar$
containing one tuple for each object consisting of the spatio-temporal data of
the object $journey$, the $licence$, the $type$, and the $model$ of the object.

In the TBA we have two relations $dataMcar$ and $dataMtrip$. $dataMcar$ contains
the static data for each object like $licence$, $type$, and $model$ together with
an object identifier $moid$. $dataMtrip$ contains for each $moid$ several tuples
each of them containing all units of a single trip of the moving object or a
single unit for a longer stop. For example each time the car drives from home node
to work node is a single trip and each time the car parks in front of the office
is also a single trip.

Besides the moving objects the \bmodb{} provides several data sets each of them
containing 100 pseudo randomly generated data objects which are used in the
benchmark queries. Table \ref{tab:queryobjects} gives an overview of this query
objects.

\begin{table}[H]
  \begin{tabularx}{|l|X|}
    \hline
    \textbf{Name of Data Set}&\textbf{Tuple Content}\\
    \hline
    $QueryPoints$&Object identifier and \dt{point} values.\\
    \hline
    $QueryRegions$&Object identifier and \dt{region} value.\\
    \hline
    $QueryInstants$&Object identifier and time stamps.\\
    \hline
    $QueryPeriods$&Object identifier and space of time.\\
    \hline
    $QueryLicences$& Object identifier and a \dt{string} representing a licence value.\\
    \hline
  \end{tabularx}
  \caption{Query Object Relations of \bmodb{}}
  \label{tab:queryobjects}
\end{table}

The \bmodb{} provides two sets of queries BerlinMOD/R and BerlinMOD/NN.
BerlinMOD/R addresses range queries and BerlinMOD/NN nearest neighbour queries.
In this paper we will focus on the range queries, which are the main aspect of the
\bmodb{} up to now.

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects:
\begin{itemize}
  \item object identity (known / unknown),
  \item dimension (standard / spatial / temporal / spatio-temporal),
  \item query interval (point / range / unbounded),
  \item condition type (single object / object relations),
  \item aggregation (with or without aggregation).
\end{itemize}
We will present the 17 queries in more detail in section \ref{sec:queries}
together with our network data model translation of this queries.
\subsection{Data Model of Free Movement in 2D space}
\label{sec:bmodbdatamod}
The data model used by the \bmodb{} is the same data model of freely moving in
two dimensional space presented in \cite{594784,335426,352963}. Spatial
positions are assumed to be located into a two dimensional space. A single spatial
position is represented by the data type \dt{point}. A \dt{point} consists of a pair of
\dt{real} values interpreted as x,y-coordinates in the assumed two dimensional plane.

Streets are represented by \dt{line} values. A \dt{line} value consists of a set
of $half segments$ representing the geometry of the line in the two dimensional space.
Each $half segment$ consists of two \dt{point} values and a boolean value telling
if the left or the right point is the dominating point of the $half segment$.

Regions are represented by the data type \dt{region}. A \dt{region} consists of
a set of $half segments$ defining the outer border of the region in
the two dimensional space. If a region contains wholes the inner border is also
formed by the $half segments$.

In \secondo{} all this spatial data types and many standard data types can be
lifted to become time dependent \dt{moving} values. For all data types \dt{$\alpha$}
the constructor \dt{moving} creates a new data type \dt{moving}(\dt{$\alpha$})
(short form \dt{m$\alpha$}).

A car may be represented by a \dt{mpoint}. A \dt{mpoint} is a \dt{point} object changing
its value within time. Therefore a \dt{mpoint} consist of a set of units called
\dt{unit}(\dt{point}) (short form \dt{upoint}). Each \dt{upoint} consists of a time
interval and two \dt{point} values. The first \dt{point} value represents the
position of the \dt{mpoint} at the start of the time interval and the second
\dt{point} value represents the position of the \dt{mpoint} at the end of the
time interval. It is assumed that the point moves on the straight line between
this two points with constant speed, which is defined by the ratio from the
distance of the two points and the length of the time interval of the unit.
All units of a \dt{mpoint} must have disjoint time intervals, because a car
cannot be at two different positions at the same time.
The units are sorted by ascending time intervals.

This spatio-temporal data model of \dt{moving} allows us to compute the position
of a \dt{mpoint} at every time instant within its definition time.
We can also compute the time instant the point passed a
given position assumed the \dt{mpoint} ever passes this position. The position of a
\dt{point} at a given time instant is represented by a \dt{intime}(\dt{point})
(short form \dt{ipoint}). A \dt{ipoint} consists of a time instant and a \dt{point} value.

Some other data types of \secondo{} which are used in the \bmodb{} are shown in
table \ref{tab:bmodbdatatypes}.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{\textwidth}{|l|X|}
\hline
\textbf{Data Type} & \textbf{Description} \\
\hline
\dt{bool} & Usual boolean data type.\\
\hline
\dt{int} & Usual integer number.\\
\hline
\dt{real} & Usual real number.\\
\hline
\dt{instant} & A point in time.\\
\hline
\dt{periods} & A set of disjoint and not connected spaces of time.\\
\hline
\dt{mbool} & A time dependent boolean value, which will be constant \true{} or \false{}
within each \dt{ubool} \\
\hline
\dt{mreal} & Time dependent real number. Each unit will be defined by a function
of time representing the \dt{real} value at each time instant.\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Other Data Types of \bmodb{}}
\label{tab:bmodbdatatypes}
\end{center}
\end{table}
\subsection{Network Data Model}
\label{sec:netdatamod}
The central idea of the network data model presented in \cite{1146465} is that
every movement is constrained by a given network and every position can be described
relative to this network. This make the data type \dt{network} to the central
data type in the network data model. All other data types of the network data model
are related to a given \dt{network} object by the unique network identifier that
is part of each \dt{network} object.

The \dt{network} object contains all spatial informations of the represented network
in three main relations, two sub relations and some indexes to support faster
access to the relations.

The relation $routes$ contains the attributes of the
streets like $id$, $route curve$, $route length$, and two boolean flags. The first
flag indicates if the route starts at the lexicographically smaller end point or not.
The second flag indicates if the lanes of the street are separated like on German
Highways or not.

The $junctions$ relation contains all attributes of the street
crossings like the two route identifiers of the first and second street meeting at the
crossing, the distance of the crossing from the start of the first respectively
second street, tuple identifiers of the both streets in the routes relation,
tuple identifiers of the sections connected by this junction in the sections relation,
and a connectivity code telling us which lanes of the two streets are connected
by the crossing.

The $sections$ relation of the \dt{network} object contains the
attributes of the street parts between two crossings or a crossing and
the end of the street. This are the route identifier of the street the section
belongs to, the tuple identifier of this street in the routes relation, the start
position and end position of the section on the street, the section curve, and
again two boolean flags with the same meaning as in the routes relation.

Furthermore there are two arrays in the \dt{network} object providing a fast
access from each section to their adjacent sections with respect to the driving
direction. Two sections are adjacent if their lanes are connected by a junction.

We created four B-Tree indexes for the route identifier attributes in the $routes$,
$junctions$ and $sections$ relation, and a R-Tree index over the curve attribute
of the $routes$ relation to support faster execution of operations dealing with
the relations of the \dt{network} object.

The data type \dt{gpoint} represents single positions in a given network. Besides the
network identifier a \dt{gpoint} consists of a route identifier, a distance from
the start of the route to the position of the \dt{gpoint} and a $side$ value
({$up$, $down$, $none$) telling us if the position is reachable from the $up$
or the $down$ side of the route in case of seperated lanes. For simple
streets or positions which are reachable from both sides of the route the side
value is always $none$.

Parts of the network, regardless if they represent paths or regions, are given as
\dt{gline} values. Besides the network identifier a \dt{gline} consists of a set
of $route interval$s, and two boolean flags. The boolean flags tell us if the
\dt{gline} is defined and if the set of $route interval$s is sorted.

Each $route interval$ consists of a route identifier identifying the route
the route interval belongs to, and the start and the end position from the
route interval on this route
\footnote{In the original paper the $route interval$ includes a $side$ value
analogous to the \dt{gpoint}. But this parameter is not part of the implementation yet.}.
We call a \label{sec:sortedgline} set of $route interval$s sorted if the following
conditions are fullfilled:
\begin{itemize}
	\item all $route interval$s are disjoint
	\item the $route interval$s are stored in ascending order of their route identifiers
	\item if two disjoint $route interval$s have the same route identifier the
$route interval$ with the smaller start position is stored first
	\item for all $route interval$s the $startPosition \le endPosition$
\end{itemize}
Many algorithms take profit from sorted \dt{gline} values. For example: If $n$
is the number of the route intervals in the \dt{gline} the
decision, if a \dt{gpoint} is inside a \dt{gline} needs O($n$) time for unsorted
and O($\log n$) time for sorted \dt{gline} values,

@@

Mostly similar to the \dt{mpoint} of the other data model there is a
\dt{mgpoint} in the network data model. A \dt{mgpoint} consists of a set of
\dt{ugpoint} with disjoint time intervals. Each \dt{ugpoint} consists of a time
interval and two \dt{gpoint} values. Every time the \dt{mgpoint} changes the route
or the speed a new \dt{ugpoint} is written. Each \dt{ugpoint} is assumed to follow
the same route from the start to the end position at the same speed. So accordingly
to the \dt{mpoint} we can compute the network position of the \dt{mgpoint} at every
time instant within the definition time of the \dt{mgpoint} as \dt{intime}(\dt{gpoint}).
